---
title: "Getting Started with maxdiff"
vignette: >
  %\VignetteIndexEntry{Getting Started with maxdiff}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## What is MaxDiff?

MaxDiff (Maximum Difference Scaling, also known as Best-Worst Scaling) is a
conjoint-style survey method where respondents repeatedly choose the **best**
and **worst** items from a small shown set. Over many tasks, these choices
reveal partworth utilities for each item on a common ratio scale.

The workhorse estimator is a hierarchical Bayesian multinomial logit (HB MNL)
model:

-   **Individual level**: each respondent has their own partworth vector
    $\beta_i$ drawn from a mixture-of-normals population distribution.
-   **Population level**: the mixture is updated across respondents via a
    Gibbs sampler (Rossi, Allenby, and McCulloch 2024).
-   **Exact joint likelihood**: each task produces *two* MNL observations —
    a K-alternative best pick and a (K-1)-alternative worst pick, jointly
    encoding the best-then-worst selection.

## Simulated example

### 1  Generate a design

```{r}
#| eval: false
library(maxdiff)

set.seed(1)
design <- make_design(
  n_items          = 8,
  n_versions       = 50,
  n_tasks          = 8,
  n_items_per_task = 4
)
head(design)
```

The output is a data frame with columns `Version`, `Task`, and `Item1`–`Item4`.

### 2  Simulate responses

```{r}
#| eval: false
# True population partworths (item 1 best, item 8 worst)
true_betas <- c(2.0, 1.5, 1.0, 0.5, 0.0, -0.5, -1.0, -2.0)

simulate_respondent <- function(version, true_b, design) {
  tasks <- design[design$Version == version, ]
  best_picks  <- integer(nrow(tasks))
  worst_picks <- integer(nrow(tasks))
  for (t in seq_len(nrow(tasks))) {
    items <- as.integer(tasks[t, paste0("Item", 1:4)])
    utils <- true_b[items] + stats::rnorm(4, sd = 1)
    best_picks[t]  <- items[which.max(utils)]
    worst_picks[t] <- items[which.min(utils)]
  }
  list(best = best_picks, worst = worst_picks)
}

n_resp   <- 200
versions <- sample(1:50, n_resp, replace = TRUE)
responses <- data.frame(
  resp_id = 1:n_resp,
  version = versions,
  matrix(0L, nrow = n_resp, ncol = 16,
         dimnames = list(NULL, c(paste0("best",  1:8),
                                  paste0("worst", 1:8))))
)

for (i in 1:n_resp) {
  r <- simulate_respondent(versions[i], true_betas, design)
  responses[i, paste0("best",  1:8)] <- r$best
  responses[i, paste0("worst", 1:8)] <- r$worst
}
```

### 3  Simple scoring (fast, no MCMC)

```{r}
#| eval: false
simple_fit <- fit_simple(
  responses         = responses,
  design            = design,
  version_col       = "version",
  best_cols         = setNames(paste0("best",  1:8), as.character(1:8)),
  worst_cols        = setNames(paste0("worst", 1:8), as.character(1:8)),
  item_labels       = paste("Item", 1:8)
)
print(simple_fit)
plot(simple_fit)
```

The simple model confirms item 1 scores highest and item 8 scores lowest.

### 4  HB MNL estimation

```{r}
#| eval: false
hb_fit <- fit_hbmnl(
  responses         = responses,
  design            = design,
  version_col       = "version",
  best_cols         = setNames(paste0("best",  1:8), as.character(1:8)),
  worst_cols        = setNames(paste0("worst", 1:8), as.character(1:8)),
  item_labels       = paste("Item", 1:8),
  R                 = 10000L,
  burnin            = 5000L,
  keep              = 5L,
  nprint            = 2000L,
  seed              = 42L
)
print(hb_fit)
```

### 5  Inspect results

```{r}
#| eval: false
# Posterior mean summary
summary(hb_fit)

# Respondent-level partworths
coef(hb_fit)

# Waterfall plot
plot(hb_fit)

# Probability-scale plot (Sawtooth-compatible)
plot(hb_fit, scale = "probability")
```

Recovered posterior means should correlate strongly with `true_betas`:

```{r}
#| eval: false
overall_means <- colMeans(hb_fit$beta_mean)
cat("Rank correlation with truth:",
    round(cor(overall_means, true_betas, method = "spearman"), 3), "\n")
```

### 6  Subgroup analysis

```{r}
#| eval: false
# Split on a simulated demographic
group_a <- responses$resp_id <=  100
group_b <- responses$resp_id  > 100

sg <- subgroup_summary(
  fit_hbmnl = hb_fit,
  subgroups = list(GroupA = group_a, GroupB = group_b)
)
plot(sg)

# Significance testing
cmp <- compare_groups(hb_fit, group_a, group_b, method = "welch")
cmp[cmp$sig_95, c("item_label", "mean_a", "mean_b", "p", "cohens_d")]
```

### 7  Publication table

```{r}
#| eval: false
# Requires the gt package
summary_table(hb_fit)
summary_table(hb_fit, scale = "probability")
summary_table(hb_fit, subgroup_results = sg)
```

## Probability scale

The **Sawtooth probability scale** transforms raw partworths to (0, 1):

$$p_j = \frac{e^{\beta_j}}{K - 1 + e^{\beta_j}}$$

An average item scores $1/K$. An item with utility $\to +\infty$ scores 1.
The scale is interpretable as "probability of being chosen best when shown
alone with K-1 average items."

```{r}
#| eval: false
to_probability_scale(colMeans(hb_fit$beta_mean), K_alts = hb_fit$K)
```

## Acknowledgments

The HB MNL Gibbs sampler in this package is a direct adaptation of
`rhierMnlRwMixture` from Peter Rossi's **bayesm** package (Rossi 2019, CRAN).
The ragged choice-set extension enabling exact best-then-worst joint likelihood
was implemented by Dan Yavorsky. All statistical credit for the sampler
algorithm goes to:

> Rossi, P. E., Allenby, G. M., & McCulloch, R. (2024).
> *Bayesian Statistics and Marketing* (2nd ed.). Wiley.
> ISBN 978-1-118-97218-3.
